---
title: "Lab 9 report (mini-project)"
author: 'Bangyan Hu (PID: A15540189)'
date: "10/26/2021"
output: pdf_document
---

#1. Exploratory data analysis
```{r}
# Save your input data file into your Project directory
fna.data <- "WisconsinCancer.csv"

# Input the data and store as wisc.df
wisc.df <- read.csv(fna.data, row.names=1)

```

```{r}
#wisc.df
```

```{r}
# We can use -1 here to remove the first column
wisc.data <- wisc.df[,-1]
```

```{r}
# Create diagnosis vector for later 
diagnosis <- factor(wisc.df[,1])
```

```{r}
View(wisc.data)
head(wisc.data)
```
```{r}
dim(wisc.data)
nrow(wisc.data)
length(wisc.data)
```
***Q1.*** How many observations are in this dataset?

There are 569 observations in this dataset.

```{r}
table(diagnosis)
```
***Q2.*** How many of the observations have a malignant diagnosis?

There are 212 observations having a malignant diagnosis.

```{r}
length(grep("_mean", colnames(wisc.data)))
length(grep("_mean", rownames(wisc.data)))
```
***Q3.*** How many variables/features in the data are suffixed with _mean?

There are 10 variables/features in the data are suffixed with _mean.

#2. Principal Component Analysis
Performing PCA
```{r}
# Check column means and standard deviations
colMeans(wisc.data)

apply(wisc.data,2,sd)
```
```{r}
# Perform PCA on wisc.data
wisc.pr <- prcomp(wisc.data, scale = TRUE)
# Look at summary of results
summary(wisc.pr)
```
***Q4.*** From your results, what proportion of the original variance is captured by the first principal components (PC1)?

From my results, 0.4427 (44.27%) of the original variance is caputured by the first principal components (PC1).

***Q5.*** How many principal components (PCs) are required to describe at least 70% of the original variance in the data?

Based on the Cumulative Proportion, 3 principal components (PCs) are required to describe at least 70% of the original variance in the data (Cumulative Proportion of PC3: 0.72636).

***Q6.*** How many principal components (PCs) are required to describe at least 90% of the original variance in the data?

Based on the Cumulative Proportion, 7 principal components (PCs) are required to describe at least 90% of the original variance in the data (Cumulative Proportion of PC7: 0.91010).

Interpreting PCA results

***Q7.*** What stands out to you about this plot? Is it easy or difficult to understand? Why?

```{r}
biplot(wisc.pr)
```
As this data contains a non-trivial number of observations and variables, this is a hot mess of a plot, which makes it is difficult to understand. We cannot clearly see the distribution pattern of observations and variables in this kind of plot. Thus, we will need to generate our own plots to make sense of this PCA result.

To make this plot ourselves we need access the PCA scores data.

```{r}
# Scatter plot observations by components 1 and 2
plot(wisc.pr$x[,1:2], col = diagnosis , 
     xlab = "PC1", ylab = "PC2")
```

***Q8.*** Generate a similar plot for principal components 1 and 3. What do you notice about these plots?
```{r}
# Repeat for components 1 and 3
plot(wisc.pr$x[,1], wisc.pr$x[,3], col = diagnosis, 
     xlab = "PC1", ylab = "PC3")
```
Because principal component 2 explains more variance in the original data than principal component 3, the first plot has a cleaner cut separating the two subgroups. For these plots, they indicate that principal component 1 is capturing a separation of malignant (red) from benign (black) samples. the principal component 2 explains less variance in the original data than principal component 1 but more variance than principal component 3. Thus, the first plot has a cleaner cut separating the two subgroups compared to the second plot (PC1 vs PC3).

```{r}
# Create a data.frame for ggplot
df <- as.data.frame(wisc.pr$x)
df$diagnosis <- diagnosis

# Load the ggplot2 package
library(ggplot2)

# Make a scatter plot colored by diagnosis
ggplot(df) + 
  aes(PC1, PC2, col=diagnosis) + 
  geom_point()
```
Variance explained
```{r}
# Calculate variance of each component
pr.var <- wisc.pr$sdev^2
head(pr.var)
```
```{r}
# Variance explained by each principal component: pve
pve <- pr.var / sum(pr.var)

# Plot variance explained for each principal component
plot(pve, xlab = "Principal Component", 
     ylab = "Proportion of Variance Explained", 
     ylim = c(0, 1), type = "o")
```
```{r}
# Alternative scree plot of the same data, note data driven y-axis
barplot(pve, ylab = "Precent of Variance Explained",
     names.arg=paste0("PC",1:length(pve)), las=2, axes = FALSE)
axis(2, at=pve, labels=round(pve,2)*100 )
```
```{r}
## ggplot based graph
#install.packages("factoextra")
library(factoextra)
fviz_eig(wisc.pr, addlabels = TRUE)
```
Communicating PCA results

***Q9.*** For the first principal component, what is the component of the loading vector (i.e. wisc.pr$rotation[,1]) for the feature concave.points_mean?

```{r}
wisc.pr$rotation["concave.points_mean", 1]
```
The component of the loading vector (i.e. wisc.pr$rotation[,1]) for the feature concave.points_mean is -0.2608538.

***Q10.*** What is the minimum number of principal components required to explain 80% of the variance of the data?

```{r}
var <- summary(wisc.pr)
var$importance[3,] < 0.8
sum(var$importance[3,] < 0.8)
var
```
Based on the Cumulative Proportion, 5 principal components (PCs) are required to describe at least 80% of the original variance in the data (Cumulative Proportion of PC5: 0.84734).

#3. Hierarchical clustering

```{r}
# Scale the wisc.data data using the "scale()" function
data.scaled <- scale(wisc.data)
```

```{r}
#Calculate the (Euclidean) distances between all pairs of observations in the new scaled dataset
data.dist <- dist(data.scaled)
```

```{r}
wisc.hclust <- hclust(data.dist, method = "complete")
wisc.hclust
```

***Q11.*** Using the plot() and abline() functions, what is the height at which the clustering model has 4 clusters?

```{r}
plot(wisc.hclust)
abline(h = 19, col="red", lty=2)
```
The height at which the clustering model has 4 clusters can be 19.

Selecting number of clusters

```{r}
wisc.hclust.clusters <- cutree(wisc.hclust, k = 4)
```

```{r}
table(wisc.hclust.clusters, diagnosis)
```
***Q12.*** Can you find a better cluster vs diagnoses match by cutting into a different number of clusters between 2 and 10?

```{r}
wisc.hclust.clusters2 <- cutree(wisc.hclust, k = 2)
wisc.hclust.clusters3 <- cutree(wisc.hclust, k = 3)
wisc.hclust.clusters4 <- cutree(wisc.hclust, k = 4)
wisc.hclust.clusters5 <- cutree(wisc.hclust, k = 5)
wisc.hclust.clusters6 <- cutree(wisc.hclust, k = 6)
wisc.hclust.clusters7 <- cutree(wisc.hclust, k = 7)
wisc.hclust.clusters8 <- cutree(wisc.hclust, k = 8)
wisc.hclust.clusters9 <- cutree(wisc.hclust, k = 9)
wisc.hclust.clusters10 <- cutree(wisc.hclust, k = 10)
table(wisc.hclust.clusters2, diagnosis)
table(wisc.hclust.clusters3, diagnosis)
table(wisc.hclust.clusters4, diagnosis)
table(wisc.hclust.clusters5, diagnosis)
table(wisc.hclust.clusters6, diagnosis)
table(wisc.hclust.clusters7, diagnosis)
table(wisc.hclust.clusters8, diagnosis)
table(wisc.hclust.clusters9, diagnosis)
table(wisc.hclust.clusters10, diagnosis)
```
As I explore how different numbers of clusters affect the ability of the hierarchical clustering to separate the different diagnoses above, by cutting into a different number of clusters between 2 and 10, I found that as the number of clusters increases, the capacity to separate the different diagnoses is higher. So, I can find a better cluster vs diagnoses match by cutting into a higher number of clusters than 4 (5 clusters can be better for a higher specificity).

Using different methods
***Q13.*** Which method gives your favorite results for the same data.dist dataset? Explain your reasoning.

The method "ward.D2" gives my favorite results for the same data.dist dataset. As the method = "ward.D2" creates groups whose variance can be minimized within clusters.
So, this has the effect of looking for spherical clusters with the process beginning with all points in individual clusters (bottom up) and then repeatedly merging a pair of clusters such that when merged there is a minimum increase in total within-cluster variance.
This process continues until a single group including all points (the top of the tree) is defined, which really makes sense for me so it would be my favorite method.

#4. OPTIONAL: K-means clustering
K-means clustering and comparing results
```{r}
wisc.km <- kmeans(scale(wisc.data), centers= 2, nstart= 20)
```

```{r}
table(wisc.km$cluster, diagnosis)
table(wisc.hclust.clusters, diagnosis)
```
***Q14.*** How well does k-means separate the two diagnoses? How does it compare to your hclust results?

```{r}
table(wisc.hclust.clusters, wisc.km$cluster)
```
Clusters 1, 2, and 4 from the hierarchical clustering model can be interpreted as the cluster 1 equivalent from the k-means algorithm, and cluster 3 can be interpreted as the cluster 2 equivalent.

Based on the table here, k-means separates the two diagnoses well (the first can be interpreted as the group largely corresponding to malignant cells and the second can be seen as the group largely corresponding to benign cells). Compared to my hclust results, k-means seems to separate the two diagnoses better as it has a higher sensitivity (same specificity) than hclust does.

#5. Combining methods

We take the results of our PCA analysis and cluster in this space 'wisc.pr$x'

```{r}
summary(wisc.pr)
```

```{r}
wisc.pr.hclust <- hclust( dist(wisc.pr$x[,1:7]), method = "ward.D2" )
```

```{r}
wisc.pr.hclust.clusters <- cutree(wisc.pr.hclust, k=2)
```

Plot my dendrogram
```{r}
plot( wisc.pr.hclust )
abline (h = 60, col = "red")
```

```{r}
grps <- cutree(wisc.pr.hclust, k=2)
table(grps)
```
```{r}
table(grps, diagnosis)
```
```{r}
plot(wisc.pr$x[,1:2], col=grps)
```
```{r}
plot(wisc.pr$x[,1:2], col=diagnosis)
```

(practice) Cut the tree into k = 2 groups

```{r}
grps <- cutree( wisc.pr.hclust, k = 2)
table(grps)
```

Cross table compare of diagnosis and my cluster groups

```{r}
table(grps, diagnosis)
```

```{r}
plot(wisc.pr$x[,1:2], col=grps)
```

```{r}
plot(wisc.pr$x[,1:2], col=diagnosis)
```

```{r}
g <- as.factor(grps)
levels(g)
```

```{r}
g <- relevel(g,2)
levels(g)
```

```{r}
# Plot using our re-ordered factor 
plot(wisc.pr$x[,1:2], col=g)
```

```{r}
library(rgl)
plot3d(wisc.pr$x[,1:3], xlab="PC 1", ylab="PC 2", zlab="PC 3", cex=1.5, size=1, type="s", col=grps)
```

```{r}
## Use the distance along the first 7 PCs for clustering i.e. wisc.pr$x[, 1:7]
wisc.pr.hclust <- hclust(dist(wisc.pr$x[,1:7]), method="ward.D2")
```

```{r}
wisc.pr.hclust.clusters <- cutree(wisc.pr.hclust, k=2)
```

***Q15.*** How well does the newly created model with four clusters separate out the two diagnoses?
```{r}
# Compare to actual diagnoses
table(wisc.hclust.clusters, diagnosis)
table(wisc.pr.hclust.clusters,diagnosis)
```
The newly created model with four clusters separates out the two diagnoses really well.

***Q16.*** How well do the k-means and hierarchical clustering models you created in previous sections (i.e. before PCA) do in terms of separating the diagnoses? Again, use the table() function to compare the output of each model (wisc.km$cluster and wisc.hclust.clusters) with the vector containing the actual diagnoses.

```{r}
table(grps, diagnosis)
```

```{r}
table(wisc.km$cluster, diagnosis)
table(wisc.hclust.clusters, diagnosis)
```
Before PCA, the k-means and hierarchical clustering models separate the diagnoses well. In the k-means model, group 1 can be interpreted as the group largely corresponding to benign cells, and group 2 can be seen as the group largely corresponding to malignant cells. In the hierarchical clustering model, group 1 can be seen as the group largely corresponding to malignant cells, and group 3 can be interpreted as the group largely corresponding to benign cells. Compared to my hclust results, k-means seems to separate the two diagnoses better as it has a higher sensitivity (same specificity) than hclust does.

#6. Sensitivity/Specificity
**Accuracy** What proportion did we get correct if we call cluster 1 M and cluster 2 B

```{r}
(329 + 188)/nrow(wisc.data)
```

**Sensitivity** refers to a test’s ability to correctly detect ill patients who do have the condition. In our example here the sensitivity is the total number of samples in the cluster identified as predominantly malignant (cancerous) divided by the total number of known malignant samples. In other words: TP/(TP+FN).

```{r}
#
188/(188+24)
```

**Specificity** relates to a test’s ability to correctly reject healthy patients without a condition. In our example specificity is the proportion of benign (not cancerous) samples in the cluster identified as predominantly benign that are known to be benign. In other words: TN/(TN+FN).

```{r}
329/(329 + 28)
```

***Q17.*** Which of your analysis procedures resulted in a clustering model with the best specificity? How about sensitivity?

```{r}
table(grps, diagnosis)
table(wisc.km$cluster, diagnosis)
table(wisc.hclust.clusters, diagnosis)
```
```{r}
#specificity
PCA_hierarchical_clustering_spe <- 329/(329+28)
kmeans_clustering_spe <- 343/(343+14)
hierarchical_clustering_spe <- 343/(343+12+2)
PCA_hierarchical_clustering_spe
kmeans_clustering_spe
hierarchical_clustering_spe
#sensitivity
PCA_hierarchical_clustering_sen <- 188/(188+24)
kmeans_clustering_sen <- 175/(175+37)
hierarchical_clustering_sen <- 172/(165+5+2+40)
PCA_hierarchical_clustering_sen
kmeans_clustering_sen
hierarchical_clustering_sen
```
The k-means and hierarchical clustering resulted in a clustering model with the best specificity, and the PCA hierarchical clustering resulted in a clustering model with the best  sensitivity.


#7. Prediction
```{r}
#url <- "new_samples.csv"
url <- "https://tinyurl.com/new-samples-CSV"
new <- read.csv(url)
npc <- predict(wisc.pr, newdata=new)
npc
```

```{r}
plot(wisc.pr$x[,1:2], col= g)
points(npc[,1], npc[,2], col="blue", pch=16, cex=3)
text(npc[,1], npc[,2], c(1,2), col="white")
```

***Q18.*** Which of these new patients should we prioritize for follow up based on your results?

New patients in 2 (Patients 2). Based on my results, since patients in 2 are in the cluster identified as predominantly malignant (cancerous), we should prioritize these new patients for follow up. 

#About this document

```{r}
sessionInfo()
```


















