---
title: "Class19 Homework (Population analysis)"
author: 'Bangyan Hu (PID: A15540189)'
date: "11/7/2021"
output: pdf_document
---
Section 4: Population Scale Analysis
[HOMEWORK]
One sample is obviously not enough to know what is happening in a population. You are
interested in assessing genetic differences on a population scale. So, you processed about
~230 samples and did the normalization on a genome level. Now, you want to find whether
there is any association of the 4 asthma-associated SNPs (rs8067378...) on ORMDL3
expression.

***Q13:*** Read this file into R and determine the sample size for each genotype and their
corresponding median expression levels for each of these genotypes.

```{r}
expr <- read.table("https://bioboot.github.io/bggn213_W19/class-material/rs8067378_ENSG00000172057.6.txt")
head(expr)
```
```{r}
nrow(expr)
```
```{r}
table(expr$geno)
```

```{r}
bp <- boxplot(exp ~ geno, data=expr,plot=F)
bp
```
The sample size for each genotype and their corresponding median expression levels for each of these genotypes:
A/A: sample size is 108, and corresponding median expression level is 31.24847.
A/G: sample size is 233, and corresponding median expression level is 25.06486.
G/G: sample size is 121, and corresponding median expression level is 20.07363.

***Q14:*** Generate a boxplot with a box per genotype, what could you infer from the relative
expression value between A/A and G/G displayed in this plot? Does the SNP effect the
expression of ORMDL3?

plot them using ggplot.

```{r}
library(ggplot2)
boxp <- ggplot(expr) + aes(geno, exp, fill=geno) + geom_boxplot(notch=TRUE)
boxp
```
Based on the boxplot demonstrated here, we can easily observe that the relative
expression value between A/A and G/G are statistically different, as the relative expression value of A/A is significantly higher than that of G/G, which indicates that the SNP effect the expression of ORMDL3.



