---
title: "Candy project Eric"
author: 'Bangyan Hu (PID: A15540189)'
date: "10/28/2021"
output: pdf_document
---

```{r}
url <- "https://raw.githubusercontent.com/fivethirtyeight/data/master/candy-power-ranking/candy-data.csv"

candy = read.csv(url, row.names=1)
head(candy, n=5)
```
***Q1.*** How many different candy types are in this dataset?
```{r}
nrow(candy)
```
There are 85 different candy types in this dataset.

***Q2.*** How many fruity candy types are in the dataset?
```{r}
sum(candy$fruity)
```
There are 38 fruity candy types are in the dataset.

```{r}
candy["Twix", ]$winpercent
```

***Q3.*** What is your favorite candy in the dataset and what is it’s winpercent value?

My favorite candy is Twix, and the winpercent value of it is 81.64291%.

***Q4.*** What is the winpercent value for “Kit Kat”?
```{r}
candy["Kit Kat", ]$winpercent
```
the winpercent value for “Kit Kat” is 76.7686.


***Q5.*** What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy["Tootsie Roll Snack Bars", ]$winpercent
```
The winpercent value for “Tootsie Roll Snack Bars” is 49.6535.

***Q6.*** Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?
```{r}
library("skimr")
skim(candy)
```
Yes, there is a variable/column that looks to be on a different scale to the majority of the other columns in the dataset -- winpercent.

***Q7.*** What do you think a zero and one represent for the candy$chocolate column?

I think that a zero represents False (no chocolate), and the one represents True (with chocolate) for the candy$chocolate column.

***Q8.*** Plot a histogram of winpercent values
```{r}
hist(candy$winpercent)
```

***Q9.*** Is the distribution of winpercent values symmetrical?

No

***Q10.*** Is the center of the distribution above or below 50%?

Below

***Q11.*** On average is chocolate candy higher or lower ranked than fruit candy?

First need to find all the chocolate candy rows in the `candy` data.
```{r}
indsc <- as.logical(candy$chocolate)
avgchocolate <- mean(candy[indsc,]$winpercent)
avgchocolate
```
```{r}
indsf <- as.logical(candy$fruit)
avgfruit <- mean(candy[indsf,]$winpercent)
avgfruit
```

```{r}
avgchocolate > avgfruit
```
Thus, on average, chocolate candy is higher ranked than fruit candy.

***Q12.*** Is this difference statistically significant?

```{r}
t.test(candy[indsc,]$winpercent, candy[indsf,]$winpercent)
```
As p-value = 2.871e-08 that is much less than 0.05, this difference is statistically significant.

***Q13.*** What are the five least liked candy types in this set?
```{r}
head(candy[order(candy$winpercent, decreasing = FALSE),], n=5)
library(dplyr)
candy %>% arrange(winpercent) %>% head(5)
```
Nik L Nip, Boston Baked Beans, Chiclets, Super Bubble, and Jawbusters.

***Q14.*** What are the top 5 all time favorite candy types out of this set?
```{r}
head(candy[order(candy$winpercent, decreasing = TRUE),], n=5)
```
ReeseÕs Peanut Butter cup, ReeseÕs Miniatures, Twix, Kit Kat, and Snickers.

***Q15.*** Make a first barplot of candy ranking based on winpercent values.

```{r}
library(ggplot2)

ggplot(candy) + 
  aes(winpercent, rownames(candy)) +
  geom_col()
```
***Q16.*** This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
library(ggplot2)

ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col()
```
Setup a color vector for coloring candy by whether it is chocolate/fruit/bar etc...
Add some useful color
```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"
```

```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols) 
```

***Q17.*** What is the worst ranked chocolate candy?

The worst ranked chocolate candy is Sixlets

***Q18.*** What is the best ranked fruity candy?

The best ranked fruity candy is Starburst.

```{r}
library(ggrepel)

# a plot of price vs win
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 5)
```

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```

***Q19.*** Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

***Q20.*** What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

***Q21.*** Make a barplot again with geom_col() this time using pricepercent and then improve this step by step, first ordering the x-axis by value and finally making a so called “dot chat” or “lollipop” chart by swapping geom_col() for geom_point() + geom_segment() (Optional).
```{r}
# a lollipop chart of pricepercent
ggplot(candy) +
  aes(pricepercent, reorder(rownames(candy), pricepercent)) +
  geom_segment(aes(yend = reorder(rownames(candy), pricepercent), 
                   xend = 0), col="gray40") +
    geom_point()
```

```{r}
library(corrplot)
cij <- cor(candy)
corrplot(cij)
```

***Q22.*** Examining this plot what two variables are anti-correlated (i.e. have minus values)?


***Q23.*** Similarly, what two variables are most positively correlated?

```{r}
pca <- prcomp(candy, scale = TRUE)
summary(pca)
```

```{r}
plot(pca$x[,1:2], col=my_cols, pch=16)
```

```{r}
# Make a new data-frame with our PCA results and candy data
my_data <- cbind(candy, pca$x[,1:3])
```

```{r}
p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)

p
```

```{r}
library(ggrepel)

p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```

```{r}
library(plotly)
ggplotly(p)
```

```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```

***Q24.*** What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

fruity, hard, and pluribus.



