---
title: "Class 13 lab report"
author: 'Bangyan Hu (PID: A15540189)'
date: "11/9/2021"
output: pdf_document
---

Read my mm-second.x.zebrafish.tsv.

```{r}
zeb <- read.delim(file="mm-second.x.zebrafish.tsv")
colnames(zeb) = c("qseqid", "sseqid", "pident", "length",
                  "mismatch", "gapopen", "qstart", "qend", 
                  "sstart", "send", "evalue", "bitscore")
head(zeb)
```

Make a histogram of the \$bitscore values.

```{r}
hist(zeb$bitscore, breaks=30)
```

Bitscores are only somewhat related to pident; they take into account not only the percent identity but the length of the alignment.

```{r}
## Asuming your blast results are stored in an object called 'b'
plot(zeb$pident  * (zeb$qend - zeb$qstart), zeb$bitscore)
```

Or using ggplot

```{r}
library(ggplot2)
ggplot(zeb, aes(pident, bitscore)) + geom_point(alpha=0.1) 
```

From this graph, we can see that bitscores are only somewhat related to pident.

Take into account not only the percent identity but the length of the alignment.

```{r}
ggplot(zeb, aes((pident * (qend - qstart)), bitscore)) + 
  geom_point(alpha=0.1) + geom_smooth()
```

There is about a straightforward relationship between bitscore and (percent identity x length).
